---
title: Pre-workshop
format: 
  gfm:
    fig-width: 7
    fig-height: 6
    wrap: none

---

Pre-workshop instructions for participants. Let's get ready to rock! ðŸš€

## R and RStudio ðŸ’»

[R](https://en.wikipedia.org/wiki/R_(programming_language)) is a fantastic software for statistical analyses. ðŸ“Š [RStudio](https://posit.co/products/open-source/rstudio/) is your trusty sidekick, helping you navigate the R universe with ease. It's like a cozy integrated development environment (IDE) for R. ðŸŒŸ

There are plenty of guides available to help you obtain or update R and RStudio. Here are a couple of them to get you started:

- [R Basics for Paleoecologists](https://ckiahtipes.github.io/) by C.A. Kiahtipes, a previous part of the APD series of workshops. 
- [Install or Update R tutorial](https://jennhuck.github.io/workshops/install_update_R.html) by Jennifer Huck. 

```{r chunk-setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Packages ðŸ“¦

Packages are like magic toolboxes ðŸ§° that contain a collection of functions for specific needs. We want to make sure that everyone has the necessary packages installed for this workshop.

### Install packages 

Let's create a list of packages that we'll need from CRAN. Here they are:

```{r pkg-list, results='hide', warning=FALSE, message=FALSE}
package_list <-
  c(
    "tidyverse", # general data wrangling and visualisation âœ¨
    "pander", # nice tables ðŸ˜
    "Bchron", # age-depth modelling ðŸ•°ï¸
    "janitor", # string cleaning ðŸ§¹
    "remotes", # installing packages from GitHub ðŸš€
    "neotoma2", # access to the Neotoma database ðŸŒ¿
    "mgcv", # GAM fitting ðŸ“ˆ
    "marginaleffects" # predicting trends ðŸ“ˆ
  )
```

Now, let's install all these amazing packages from CRAN:

```{r pkg-install, eval = FALSE}
lapply(
  package_list, utils::install.packages
)
```

#### Install the REcopol package ðŸŒ¼

The {REcopol} package provides easy-to-use functions for analyzing fossil pollen data. Take a peek at its [website](https://hope-uib-bio.github.io/R-Ecopol-package/) for more information.

âš ï¸ However, please note that there's a small chance it won't work on the newest version of R or on operating systems other than Windows. âš ï¸

Let's try installing the package from GitHub:

```{r pkg-install-gh, eval = FALSE}
# Install R-Ecopol
remotes::install_github("HOPE-UIB-BIO/R-Ecopol-package")
```

In case the installation doesn't go as planned, don't worry! I've got a nifty workaround prepared for you. You can use it during the workshop. Just check if you really need it (by running the following code) and let me know during the workshop. ðŸ˜‰

```{r need-workaround}
if (
  "REcopol" %in% utils::installed.packages()
) {
  cat("REcopol package is already installed")
} else {
  warning("You need to use workabout during the workshop")
}
```


### Test if everything is set up âœ…

Let's do a quick test to make sure everything is in order. Running the following code should produce `"Everything is good to go"` instead of an error message saying `"All required packages are not installed"`.

```{r pkg-test, eval = FALSE}
if (
  isTRUE(
    all(
      c(package_list) %in%
        as.data.frame(
          utils::installed.packages()
        )[, 1]
    )
  )
) {
  cat("Everything is good to go")
} else {
  warning("All required packages are not installed")
}
```